`r params$project_title`

### 2020-11-27 reports/rmarkdown/03_integrated_datasets.Rmd
updates integrated report

```{r, echo = FALSE}
library(igraph)
library(rafalib)
library(fgsea)
library(scales)

remotes::install_github("czarnewski/niceRplots",upgrade = "always",force = T)
library(niceRplots)
```

@@

#Graph construction  
fct <- .25 # FC threshold for differential expression
pvt <- 0.01 # Pvalue threshold for differential expression
min_pct <- .1 # minimun level of detected bacteria in each sample group

@@

@@

```{r, echo = FALSE,fig.width=10,fig.height=14}
res <- data.frame( matrix(0,nrow = 1,ncol = 6) )
for(i in levels(datasets)){
  for(j in rownames(all_microbiome) ){

@@

image(t(datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]]
        [ names(sort(ord)),sample_order]),
      col=heat_pal,axes=F,border=NA,
      main="CVL week 2",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
        rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V2)), pch=16,
        col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)
points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
        rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V2)), pch=16,
        col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)
```

**Figure 5.** Comparisson of microbiome datasets, showing only the significant bacteria. Samples are ordered by the CVL3 groupings. The colors represent their respective bacterial groupings for each dataset.

@@

```{r, echo = FALSE,fig.width=8,fig.height=8}
metadata_parameters <- c("Antibiotics","Antibiotics_v2","Antibiotics_v3",
                         "Bacterio_KOH_v2","Bacterio_KOH_v3","Bacterio_TV_v3",
                         "BV_Clue_Cells_v2","BV_Clue_Cells_v3",
                         "BV_Diagnosis_v2","BV_Diagnosis_v3",
                         "BV_Lactobacillus_v2","BV_Lactobacillus_v3",
                         "BV_Monbilicus_v2","BV_Monbilicus_v3",
                         "BV_Vaginal_Garda_v2","BV_Vaginal_Garda_v3",
                         "blood_cervix_v2","blood_cervix_v3",
                         "cervical_inflammation_v2","cervical_inflammation_v3",
                         "HIVstatus","Contraception",
                         "CT_v2","CT_v3","regular_partner",
                         "vaginal_douching_v2","vaginal_douching_v3",
                         "louvain_tissue_V3","louvain_CVL_V2","louvain_CVL_V3",
                         "group_M9_CVLv3","group_M6_CVLv3","group_M4_CVLv3")

@@

clean_Ftest <- Ftest
# clean_Ftest <- ( clean_Ftest + t(clean_Ftest) ) / 2
clean_Ftest <- -log10(clean_Ftest)
clean_Ftest[clean_Ftest > 4] <- 4
# clean_Ftest[ clean_Ftest > 0.05 ] <- 1

@@

```{r, echo = FALSE,fig.width=10,fig.height=13}
myletter <- letters

@@

@@

```{r, echo = FALSE,fig.width=10,fig.height=10}
cors <- cor(t(all_microbiome[rowSums(all_microbiome > 0)>10,]))

#Graph construction  
k <- 10 # k for KNN
P <- 0.3 # prunning threshold of SNN

@@

hist(pSNN[pSNN>0],breaks = seq(0,1,length.out = 50),
     border=NA,col="grey60",las=1,
     xaxs="i",yaxs="i",
     main="SNN jaccard",xlim=c(0,1))
abline(v=P,col="red",lty=2,xpd=F)

hist(cors[pSNN>0],breaks = seq(-1,1,length.out = 50),
     border=NA,col="grey60",las=1,
     xaxs="i",yaxs="i",
     main="correlation coeficient",xlim=c(-1,1))
abline(v=0,col="red",lty=1,xpd=F)

@@

```{r, echo = FALSE,fig.width  = 10, fig.height=3}
mypar(2,2,mar=c(0,0,2,10))

set.seed(1)
l <- layout_with_fr(g,niter=3000,start.temp=30)

cl <- igraph::cluster_louvain(g)
table(cl$membership)

@@

```{r, echo = FALSE,fig.width  = 10, fig.height=10}
mypar(2,5,mar=c(1,1,2,1))
for(j in sort(unique(cl$membership)) ){
  empty_plot()
  text( 0,1,labels = paste0(rownames(cors)[ cl$membership == j ],collapse = "\n"),adj=c(0,1),cex=.7)
  title(main=paste0("community",j))
}
```

@@

mypar(3,1,mar=c(1,2,1,0.5))
sample_order <- order(df$louvain_tissue_V3)
image(t(module_means_per_dataset[["ASV_tissue_V3_normalized_batch_corrected"]]
        [ ,sample_order]),
      col=heat_pal,axes=F,border=NA,
      main="tissue week 3",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
points( (0:(length(df$louvain_tissue_V3)-1) )/(length(df$louvain_tissue_V3)-1),
        rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_tissue_V3)), pch=16,
        col=pal[factor(df$louvain_tissue_V3)][sample_order] ,xpd=T)
points( (0:(length(df$louvain_tissue_V3)-1) )/(length(df$louvain_tissue_V3)-1),
        rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_tissue_V3)), pch=16,
        col=pal[factor(df$louvain_tissue_V3)][sample_order] ,xpd=T)
text( rep(par("usr")[c(1)]-.01*diff(par("usr")[c(1,2)]), length(sort(unique(cl$membership))) ),
         (0:(length( sort(unique(cl$membership)) )-1) )/(length(sort(unique(cl$membership)))-1),
         labels= rev(sort(unique(cl$membership)) )    )

sample_order <- order(df$louvain_CVL_V3)
image(t(module_means_per_dataset[["ASV_CVL_V3_normalized_batch_corrected"]]
        [ ,sample_order]),
      col=heat_pal,axes=F,border=NA,
      main="CVL week 3",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
points( (0:(length(df$louvain_CVL_V3)-1) )/(length(df$louvain_CVL_V3)-1),
        rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V3)), pch=16,
        col=pal[factor(df$louvain_CVL_V3)][sample_order] ,xpd=T)
points( (0:(length(df$louvain_CVL_V3)-1) )/(length(df$louvain_CVL_V3)-1),
        rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V3)), pch=16,
        col=pal[factor(df$louvain_CVL_V3)][sample_order] ,xpd=T)
text( rep(par("usr")[c(1)]-.01*diff(par("usr")[c(1,2)]), length(sort(unique(cl$membership))) ),
         (0:(length( sort(unique(cl$membership)) )-1) )/(length(sort(unique(cl$membership)))-1),
         labels= rev(sort(unique(cl$membership)) )    )

sample_order <- order(df$louvain_CVL_V2)
image(t(module_means_per_dataset[["ASV_CVL_V2_normalized_NOT_batch_corrected"]]
        [ ,sample_order]),
      col=heat_pal,axes=F,border=NA,
      main="CVL week 2",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
        rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V2)), pch=16,
        col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)
points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
        rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V2)), pch=16,
        col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)
text( rep(par("usr")[c(1)]-.01*diff(par("usr")[c(1,2)]), length(sort(unique(cl$membership))) ),
         (0:(length( sort(unique(cl$membership)) )-1) )/(length(sort(unique(cl$membership)))-1),
         labels= rev(sort(unique(cl$membership)) )    )
```

**Figure 5.** Comparisson of microbiome datasets, showing only the distribution of different bacterial communities (1-8). Samples are ordered by the CVL3 groupings. The colors represent their respective bacterial groupings for each dataset.

```{r, echo = FALSE,fig.width=10,fig.height=13}
myletter <- letters

@@

  # mypar(4,3,mar=c(4,2,2,2))
  #filter TRX dataset
  TRX <- TRX [ , samples_TRX_and_CVL3 ]
  TRX <- TRX [ rowSums(TRX>0)>= 2 , ]
  top_vars_TRX <- names(sort(apply(TRX,1,var),decreasing = T)[1:2000] )
  # plot(apply(TRX,1,mean),apply(TRX,1,var),col=ifelse(rownames(TRX) %in% top_vars_TRX,"red","grey20"),cex=.2)
  TRX <- TRX[ top_vars_TRX , ]
  dim(TRX)

@@

@@

  # pvalues[ pvalues < -log10(0.05)] <- 0
  # sort(rowSums(pvalues>0),T)
  # sort(colSums(pvalues>0),T)
  NESes <- lapply(enrichments,function(x) setNames(x$NES,x$pathway) )
  NESes <- t(as.data.frame(NESes))
  NESes[is.na(NESes)] <- 0
  NESes[ pvalues < -log10(0.1) ] <- 0
  NESes <- NESes[rowSums(NESes!=0) >= 2 , colSums(NESes!=0) >= 2]

>>>>

```{r, echo = FALSE}
library(igraph)
library(rafalib)
library(fgsea)
library(scales)
remotes::install_github("czarnewski/niceRplots",upgrade = "always",force = T)
library(niceRplots)
```

@@

#Graph construction  
fct <- .25 # FC threshold for differential expression
pvt <- 0.01 # Pvalue threshold for differential expression
min_pct <- .05 # minimun level of detected bacteria in each sample group

@@

```{r, echo = FALSE,fig.width=10,fig.height=9}
all_microbiome <- rbind(datasets_all_samples[["ASV_tissue_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])
all_microbiome <- all_microbiome[ rowSums(all_microbiome>0)>=2 ,  ]

zscore <- t(apply(all_microbiome,1,function(i){scale(i,T,T)}))
zscore[ zscore > 6 ] <- 6
zscore[ zscore < -6 ] <- -6
rownames(zscore) <- rownames(all_microbiome); colnames(zscore) <- colnames(all_microbiome)

#Compute sample-wise correlations
cors <- cor(zscore, method = "pearson")

#Graph construction  
k <- 5 # k for KNN
P <- 0.4# prunning threshold of SNN

#Compute KNN graph from correlation matrix
knn <- RANN::nn2(cors,k = k)
# knn <- RANN::nn2(t(zscore),k = k)
nn <- matrix(0, nrow(knn$nn.idx), nrow(knn$nn.idx))
for(idx in 1:nrow(knn$nn.idx) ){  nn[idx,knn$nn.idx[idx,]] <- 1 }

# hist(pSNN[pSNN>0],breaks = seq(0,1,length.out = 50),
#      border=NA,col="grey60",las=1,
#      xaxs="i",yaxs="i",
#      main="SNN jaccard",xlim=c(0,1))
# abline(v=P,col="red",lty=2,xpd=F)
# 
# hist(cors[pSNN>0],breaks = seq(-1,1,length.out = 50),
#      border=NA,col="grey60",las=1,
#      xaxs="i",yaxs="i",
#      main="correlation coeficient",xlim=c(-1,1))
# abline(v=0,col="red",lty=1,xpd=F)

g <- graph_from_adjacency_matrix(pSNN,mode = "undirected",diag = F,weighted = T)

mypar(2,2,mar=c(0,0,2,10))

set.seed(1)
# l <- layout_with_graphopt(g,niter = 1000)
l <- layout_with_fr(g,niter=3000,start.temp=30)

cl <- igraph::cluster_louvain(g)
# cl <- igraph::walktrap.community(g)
metadata$joint_clustering <- cl$membership

plot( g , vertex.label.cex=0.000001 , vertex.color = pal[factor(cl$membership)] , 
      edge.width=  ( E(g)$weight / max(E(g)$weight)) ,layout=l,
      vertex.size=10,
      edge.color=colorRampPalette(c("grey95","black"))(90) [ round( E(g)$weight / max(E(g)$weight) * 88 )+1 ] )
legend(x = par("usr")[c(2)]+.2*diff(par("usr")[c(1,2)]),
       y = par("usr")[c(4)],
       legend = levels(factor(cl$membership)),
       bty = "n",pch = 21,pt.bg = pal,pt.cex = 1,xpd=T)
title("Louvain clusters",cex.main=1,font.main=1)

**Figure 1b.** Sample embedding of `r k`-SNN graph clustered using Louvain based on the normalized bacterial abundances across all microbiome datasets (including samples not present in all datasets).

```{r, echo = FALSE,fig.width=10,fig.height=4}
myletters <- letters

mypar(1,2,mar=c(2,15,1,1))

for (m in c("ASV_tissue_V3_normalized_batch_corrected",
            "ASV_CVL_V3_normalized_batch_corrected",
            "ASV_CVL_V2_normalized_NOT_batch_corrected")){

  all_microbiome <- datasets_all_samples[[m]]

  datasets <- factor(metadata$joint_clustering)
  # datasets <- datasets[ colSums(all_microbiome)>0 ]
  # all_microbiome <- all_microbiome[,colSums(all_microbiome)>0]

      set.seed(1)
      a <- c(sample(all_microbiome[j,datasets == i],NN))
      set.seed(1)
      b <- sample(all_microbiome[j,datasets != i],NN)

  res$fc <- as.numeric(res$fc)
  res$FDR <- p.adjust(res$pvalue)
  # res <- res[order(res$pvalue),]
  res <- res[abs(res$fc) > fct,]
  res <- res[abs(res$pvalue) < 0.05,]
  res <- res[ (res$perc.1 > min_pct) | (res$perc.2 > min_pct) ,]
  dim(res)

  plot_dots( all_microbiome , names(sort(ord)) , clustering = datasets,show_grid = T,cex.main=1,font.main=1,
            cex.row = .8,cex.col = .8,srt = 0,main = m)
  add_letter(myletters); myletters <- myletters[-1]

  barlist( all_microbiome , names(sort(ord)) , clustering = datasets,show_grid = T,cex.main=1,font.main=1,
            cex.axis = .8,srt = 0,main = m)
  add_letter(myletters); myletters <- myletters[-1]

**Figure 2.** Differential bacterial abundance across joint clustered samples and compared across microbiome datasets. The results are shown both as **a)** Dot plots and **b)** barplots Bacteria with log2FC above `r fct` and p-value below `r pvt` were considered significant and were sorted by the highest expression.

@@

```{r, echo = FALSE,fig.width=10,fig.height=12}
res <- data.frame( matrix(0,nrow = 1,ncol = 6) )
for(i in levels(datasets)){
  for(j in rownames(all_microbiome) ){

@@

image(t(datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]]
        [ names(sort(ord)),sample_order]),
      col=heat_pal,axes=F,border=NA,
      main="CVL week 2",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
        rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V2)), pch=16,
        col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)
points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
        rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
            length(df$louvain_CVL_V2)), pch=16,
        col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)

# **Figure 5.** Comparisson of microbiome datasets, showing only the significant bacteria. Samples are ordered by the CVL3 groupings. The colors represent their respective bacterial groupings for each dataset.

@@

```{r, echo = FALSE,fig.width=8,fig.height=8}
metadata_parameters <- c("Antibiotics","Antibiotics_v2","Antibiotics_v3",
                         "Bacterio_KOH_v2","Bacterio_KOH_v3","Bacterio_TV_v3",
                         "BV_Clue_Cells_v2","BV_Clue_Cells_v3",
                         "BV_Diagnosis_v2","BV_Diagnosis_v3",
                         "BV_Lactobacillus_v2","BV_Lactobacillus_v3",
                         "BV_Monbilicus_v2","BV_Monbilicus_v3",
                         "BV_Vaginal_Garda_v2","BV_Vaginal_Garda_v3",
                         "blood_cervix_v2","blood_cervix_v3",
                         "cervical_inflammation_v2","cervical_inflammation_v3",
                         "HIVstatus","Contraception",
                         "CT_v2","CT_v3","regular_partner",
                         "vaginal_douching_v2","vaginal_douching_v3",
                         "louvain_tissue_V3","louvain_CVL_V2","louvain_CVL_V3",
                         "group_M9_CVLv3","group_M6_CVLv3","group_M4_CVLv3","joint_clustering")

@@

clean_Ftest <- Ftest
# clean_Ftest <- ( clean_Ftest + t(clean_Ftest) ) / 2
clean_Ftest <- -log10(clean_Ftest)
clean_Ftest[clean_Ftest > 5] <- 5
# clean_Ftest[ clean_Ftest > 0.05 ] <- 1

@@

```{r, eval=F, echo = FALSE,fig.width=10,fig.height=13}
myletter <- letters

@@

}

# **Figure 9.** Heatmaps showing the functional association of microbiome datasets with the expression profiles in the RNAseq dataset. Briefly, bacterial abundances from each dataset were correlated with the gene expression of the top 5000 highly variable genes from the RNAseq dataset, generating a correlation matrix between bacteria and genes. Then for each bacteria, we rank genes based on their correlation to that bacteria and perform gene set enrichment anlaysis (GSEA) using the KEGG gene annotation database. This, in turn, will result in a matrix associating every bacteria with every KEGG process in the tissue. The heatmap shows the normalized enrichment score (NES). Only enrichments with pvalue below 0.05 are shown. Bacteria and pathways significant in less that 10 pathways and bacteria, respectivelly, were omitted.

@@

```{r, echo = FALSE,fig.width=10,fig.height=10}

all_microbiome <- cbind(datasets_all_samples[["ASV_tissue_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V3_normalized_batch_corrected"]],
                        datasets_all_samples[["ASV_CVL_V2_normalized_NOT_batch_corrected"]])

cors <- cor(t(all_microbiome[rowSums(all_microbiome > 0)>=10,]))

#Graph construction  
k <- 5 # k for KNN
P <- 0.2 # prunning threshold of SNN

@@

# hist(pSNN[pSNN>0],breaks = seq(0,1,length.out = 50),
#      border=NA,col="grey60",las=1,
#      xaxs="i",yaxs="i",
#      main="SNN jaccard",xlim=c(0,1))
# abline(v=P,col="red",lty=2,xpd=F)
# 
# hist(cors[pSNN>0],breaks = seq(-1,1,length.out = 50),
#      border=NA,col="grey60",las=1,
#      xaxs="i",yaxs="i",
#      main="correlation coeficient",xlim=c(-1,1))
# abline(v=0,col="red",lty=1,xpd=F)

@@

```{r, echo = FALSE,fig.width=10,fig.height=9}
mypar(2,2,mar=c(0,0,2,10))

set.seed(1)
l <- layout_with_fr(g,niter=3000,start.temp=1)

# cl <- igraph::cluster_louvain(g)
cl <- igraph::walktrap.community(g)

table(cl$membership)

@@

**Figure 9a.** Bacterial community embedding of `r k`-SNN graph clustered using walktrap community detection algorithm based on the normalized bacterial abundances across all microbiome datasets (including samples not present in all datasets).

```{r, echo = FALSE,fig.width  = 10, fig.height=12}
mypar(3,6,mar=c(1,1,2,1))
for(j in sort(unique(cl$membership)) ){
  empty_plot()
  text( 0,1,labels = paste0(rownames(cors)[ cl$membership == j ],collapse = "\n"),adj=c(0,1),cex=.7)
  title(main=paste0("community",j))
}
```

**Figure 9b.** List of bacterial belongin to each community.

```{r, echo = FALSE,fig.width  = 10, fig.height=5}
taxa <- read.csv2("../../results/taxonomy.csv",row.names = 1,stringsAsFactors = F,as.is = T)
taxa[is.na(taxa)] <- "NA"
rownames(cors) %in% rownames(taxa)

mypar(2,2,mar=c(1,10,2,10))
x <- taxa[match(rownames(cors),rownames(taxa)),2]
y <- taxa[match(rownames(cors),rownames(taxa)),3]
plot_sankey( data.frame( x[y!="NA"] , y[y!="NA"]), color_by = 1,
       plot_labels = T, plot_weights = F,order_1_by = "total",
       order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
       pal=hue_pal()(length(unique(x))))

plot_sankey( data.frame( y[y!="NA"] , cl$membership[y!="NA"]), color_by = 1,
     plot_labels = T, plot_weights = F,order_1_by = "total",
     order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
     pal=hue_pal()(length(unique(y))))

plot_sankey( data.frame( x[y!="NA"] , cl$membership[y!="NA"]), color_by = 1,
     plot_labels = T, plot_weights = F,order_1_by = "total",
     order_2_by = "total",use_w1 = T,use_w2 = T,gapv = .02,gap2v = 0,smoothing = .1,
     pal=hue_pal()(length(unique(x))))

**Figure 9b.** Taxonomic annotation of bacterial communities.

@@

# mypar(3,1,mar=c(1,2,1,3))
# sample_order <- order(df$louvain_tissue_V3)
# image(t(module_means_per_dataset[["ASV_tissue_V3_normalized_batch_corrected"]]
#         [ ,sample_order]),
#       col=heat_pal,axes=F,border=NA,
#       main="tissue week 3",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
# points( (0:(length(df$louvain_tissue_V3)-1) )/(length(df$louvain_tissue_V3)-1),
#         rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
#             length(df$louvain_tissue_V3)), pch=16,
#         col=pal[factor(df$louvain_tissue_V3)][sample_order] ,xpd=T)
# points( (0:(length(df$louvain_tissue_V3)-1) )/(length(df$louvain_tissue_V3)-1),
#         rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
#             length(df$louvain_tissue_V3)), pch=16,
#         col=pal[factor(df$louvain_tissue_V3)][sample_order] ,xpd=T)
# text(  par("usr")[c(2)] , seq(1,0,length.out = nrow(NESes)),
#         labels = rownames(NESes), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)
# 
# 
# sample_order <- order(df$louvain_CVL_V3)
# image(t(module_means_per_dataset[["ASV_CVL_V3_normalized_batch_corrected"]]
#         [ ,sample_order]),
#       col=heat_pal,axes=F,border=NA,
#       main="CVL week 3",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
# points( (0:(length(df$louvain_CVL_V3)-1) )/(length(df$louvain_CVL_V3)-1),
#         rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
#             length(df$louvain_CVL_V3)), pch=16,
#         col=pal[factor(df$louvain_CVL_V3)][sample_order] ,xpd=T)
# points( (0:(length(df$louvain_CVL_V3)-1) )/(length(df$louvain_CVL_V3)-1),
#         rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
#             length(df$louvain_CVL_V3)), pch=16,
#         col=pal[factor(df$louvain_CVL_V3)][sample_order] ,xpd=T)
# text(  par("usr")[c(2)] , seq(1,0,length.out = nrow(NESes)),
#         labels = rownames(NESes), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)
# 
# 
# sample_order <- order(df$louvain_CVL_V2)
# image(t(module_means_per_dataset[["ASV_CVL_V2_normalized_NOT_batch_corrected"]]
#         [ ,sample_order]),
#       col=heat_pal,axes=F,border=NA,
#       main="CVL week 2",xlab=" ",ylab="",line=.4,cex.main=1,font.main=1)
# points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
#         rep(par("usr")[c(3)]-.01*diff(par("usr")[c(3,4)]),
#             length(df$louvain_CVL_V2)), pch=16,
#         col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)
# points( (0:(length(df$louvain_CVL_V2)-1) )/(length(df$louvain_CVL_V2)-1),
#         rep(par("usr")[c(4)]+.01*diff(par("usr")[c(3,4)]),
#             length(df$louvain_CVL_V2)), pch=16,
#         col=pal[factor(df$louvain_CVL_V2)][sample_order] ,xpd=T)
# text(  par("usr")[c(2)] , seq(1,0,length.out = nrow(NESes)),
#         labels = rownames(NESes), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)

mypar(3,3,mar=c(3,5,2,2))

for( i in c("louvain_tissue_V3","louvain_CVL_V3","louvain_CVL_V2")){
  for( j in c("ASV_tissue_V3_normalized_batch_corrected",
          "ASV_CVL_V3_normalized_batch_corrected",
          "ASV_CVL_V2_normalized_NOT_batch_corrected")){
    temp_group <- as.character(df[,i])
    temp_group[ is.na(temp_group) ] <- "NA"
    temp_group[ temp_group == "" ] <- "NA"
    temp_group <- factor(temp_group)

    barlist(data=module_means_per_dataset[[j]],
            genes=rownames(module_means_per_dataset[[j]]),
            labels = paste0("bac.com.",rownames(module_means_per_dataset[[j]])),
            clustering=temp_group, main=paste0(i,"\n", sub("_norm.*","",j)),srt=45)
  }
}

```{r, eval=T, echo = FALSE,fig.width=10, fig.height=14}
mypar(7,3,mar=c(3,5,2,2))

for( i in c("louvain_tissue_V3","louvain_CVL_V3","louvain_CVL_V2")){
  for( j in c("ASV_tissue_V3_normalized_batch_corrected",
          "ASV_CVL_V3_normalized_batch_corrected",
          "ASV_CVL_V2_normalized_NOT_batch_corrected")){
    temp_group <- as.character(df[,i])
    temp_group[ is.na(temp_group) ] <- "NA"
    temp_group[ temp_group == "" ] <- "NA"
    temp_group <- factor(temp_group)

    barlist(data=module_means_per_dataset[[j]],
            genes=rownames(module_means_per_dataset[[j]]),
            labels = paste0("bac.com.",rownames(module_means_per_dataset[[j]])),
            clustering=temp_group, main=paste0(i,"\n", sub("_norm.*","",j)),srt=45)
  }
}

for( i in c("group_M4_CVLv3","group_M6_CVLv3","group_M9_CVLv3","joint_clustering")){
  for( j in c("ASV_tissue_V3_normalized_batch_corrected",
          "ASV_CVL_V3_normalized_batch_corrected",
          "ASV_CVL_V2_normalized_NOT_batch_corrected")){

    temp_group <- as.character(metadata[,i])
    temp_group[ is.na(temp_group) ] <- "NA"
    temp_group[ temp_group == "" ] <- "NA"
    temp_group <- factor(temp_group)

    barlist(data=module_means_per_dataset[[j]],
            genes=rownames(module_means_per_dataset[[j]]),
            labels = paste0("bac.com.",rownames(module_means_per_dataset[[j]])),
            clustering=temp_group, main=paste0(i,"\n", sub("_norm.*","",j)),srt=45)
  }
}
```

**Figure 5.** Comparisson of microbiome communities across all datasets and all clustering methods.

```{r, echo = FALSE,fig.width=10,fig.height=8}
myletter <- letters

NESse_list <- list()

for(i in c("ASV_tissue_V3_normalized_batch_corrected",
           "ASV_CVL_V3_normalized_batch_corrected")){

  cat(paste0("\nCorrelation between RNAseq and ",sub("_norm.*","",i)))

@@

@@

  bacteria_use <- colnames(cors)
  gmt <- gmtPathways("../../supplementary_files/h.all.v6.2.symbols.gmt.txt")
  cl <- parallel::makePSOCKcluster(parallel::detectCores()-1)
  invisible(parallel::clusterEvalQ(cl, {c("gmt","cors");library(fgsea);library(stats);library(base)}))
  enrichments <- parallel::parLapply(cl,bacteria_use,gmt=gmt,cors=cors,function(x,gmt,cors){
    res <- fgsea(pathways = gmt,stats = sort(cors[,x],decreasing = T),nper=10000)
    return(res)
  })
  parallel::stopCluster(cl)
  names(enrichments) <- paste0( "bac.com.", bacteria_use )

@@

  # pvalues[ pvalues < -log10(0.05)] <- 0
  # sort(rowSums(pvalues>0),T)
  # sort(colSums(pvalues>0),T)
  NESes <- lapply(enrichments,function(x) setNames(x$NES,x$pathway) )
  NESes <- t(as.data.frame(NESes))
  NESes[is.na(NESes)] <- 0
  NESes[ pvalues < -log10(0.05) ] <- 0
  # NESes <- NESes[rowSums(NESes!=0) >= 2 , colSums(NESes!=0) >= 2]

  # o_kegg <- hclust( as.dist( (1-cor(NESes))/2 ),"ward.D2")$order
  # o_bacs <- hclust( as.dist( (1-cor(t(NESes)))/2 ),"ward.D2")$order
  # o_kegg <- hclust( as.dist( (1-cor(NESes))/2 ),"ward.D2")$order
  # o_bacs <- hclust( as.dist( (1-cor(t(NESes) ))/2 ),"ward.D2")$order
  # NESes <- NESes[o_bacs,o_kegg]

  NESse_list[[i]] <- NESes

  # image( t(NESes[nrow(NESes):1,]),col=colorRampPalette(c("navy","grey95","firebrick"))(91),breaks=seq(-3,3,length.out = 92),
  #        axes=F,border=NA,main=i,xlab="",ylab="",line=.4,cex.main=1,font.main=1)
  # text(  par("usr")[c(4)] , seq(1,0,length.out = nrow(NESes)),
  #       labels = rownames(NESes), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)
  # text( seq(0,1,length.out = ncol(NESes)) , par("usr")[c(1)],
  #       labels = colnames(NESes), srt = 90, adj = c(1,.5), xpd = TRUE, cex=.4)
  # add_letter(myletter); myletter <- myletter[-1]
  # empty_plot()
  # cat("\\clearpage")

}

common_pathways <- unique( unlist( lapply( NESse_list , function(x) colnames(x)[colSums(x!=0)>=1] ) ) )

layout(matrix(c(1,2,3,4,4),nrow = 1))
par(mar=c(6,1,2,0))
for( i in names(NESse_list)){
  x <- NESse_list[[i]][,common_pathways]
  image( x[,ncol(x):1],col=colorRampPalette(c("navy","grey95","firebrick"))(91),breaks=seq(-5,5,length.out = 92),
         axes=F,border=NA,main=sub("_norm.*","",i),xlab="",ylab="",line=.4,cex.main=1,font.main=1)
  # add_letter(myletter); myletter <- myletter[-1]
  text(   seq(1,0,length.out = nrow(x)) , par("usr")[c(3)] ,
      labels = rownames(x), srt = 90, adj = c(1,.5), xpd = TRUE, cex=1)
}
empty_plot(xaxs="i",yaxs="i",
           xlim=par("usr")[c(1,2)],ylim=par("usr")[c(3,4)])
text(  par("usr")[c(1)],seq(0,1,length.out = ncol(x)) ,
      labels = sub("HALLMARK_","",colnames(x)), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)

cat("\\clearpage")

#Combined representation of the two datasets
layout(matrix(c(1,2,2,2,2),nrow = 1))
par(mar=c(6,1,3,0))
x <- (NESse_list[[1]][,common_pathways] + NESse_list[[2]][,common_pathways]) / 2
x[NESse_list[[1]][,common_pathways] == 0] <- 0
x[NESse_list[[2]][,common_pathways] == 0] <- 0
x <- x[, colSums(x!=0)>0]
image( x[,ncol(x):1],col=colorRampPalette(c("navy","grey95","firebrick"))(91),breaks=seq(-5,5,length.out = 92),
       axes=F,border=NA,main="CVL_v3 + tissue_v3\ncombined",xlab="",ylab="",line=.4,cex.main=1,font.main=1)
# add_letter(myletter); myletter <- myletter[-1]
text(   seq(0,1,length.out = nrow(x)) , par("usr")[c(3)] ,
    labels = rownames(x), srt = 90, adj = c(1,.5), xpd = TRUE, cex=1)
empty_plot(xaxs="i",yaxs="i",
           xlim=par("usr")[c(1,2)],ylim=par("usr")[c(3,4)],main="HALLMARK")
text(  par("usr")[c(1)],seq(0,1,length.out = ncol(x)) ,
      labels = sub("HALLMARK_","",colnames(x)), srt = 0, adj = c(0,.5), xpd = TRUE, cex=1)

#Combined representation of the two datasets
layout(matrix(c(1,1,1,1,1,2,2),nrow = 1))
par(mar=c(6,30,3,0))
x <- (NESse_list[[1]][,common_pathways] + NESse_list[[2]][,common_pathways]) / 2
x[NESse_list[[1]][,common_pathways] == 0] <- 0
x[NESse_list[[2]][,common_pathways] == 0] <- 0
x <- x[, colSums(x!=0)>0]

TRX <- datasets_all_samples[["Tissue_RNAseq_V3_normalized"]]
non_zeros <- colSums(TRX)!=0
TRX <- TRX [ top_vars_TRX , non_zeros ]
pathway_means <- t(sapply( colnames(x) , function(i){
  i <- gmt[[i]]
  i <- as.matrix( TRX[ rownames(TRX) %in% i , ] )
  if(ncol(i)>1){
    i <- ( i - apply(i,1,min) ) / ( apply(i,1,max)- apply(i,1,min) )
    i <- colMeans(i)
  } else {
    i <- (i - min(i)) / (max(i) - min(i))
  }
  return( i - min(i) )
} ))
plot_dots(data=pathway_means,main="patient groups",cex.main=1,font.main=1,
        genes=rownames(pathway_means),
        clustering=metadata$joint_clustering[non_zeros],
        srt=0,pal=c("blue","blue","navy","navy","grey95","grey95","firebrick3","black"))
